Do we need detailed demographic data to forecast plant population responses to climate change?

### 2016-07-06 manuscript/supporting_information.Rmd
updates stats notation as per MBH edits

##  Hierarchical Bayesian models for vital rates
Here we describe the statistical models in complete detail, including full description of the Bayesian models and Stan code for each model. We write each model as three parts: (1) a data model (likelihood), (2) a process model, and (3) a parameter model. We then combine these three models to write the posterior and joint distributions as a single model statement. Following the full model expression we provide the Stan code that directly corresponds the statement of posterior and joint distributions. In our expressions of the posterior and joint distributions we use general probability notation where [.] represents some unspecified distribution. We do this for clarity of model presentation. All prior distributions are defined in the parameter model sections. For clarity, we avoid subscripting species.

@@

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\beta_{0,t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
| $\beta_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\beta_{s,t}$ | size effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_s}$ | global size effect | ``b1_mu`` |
| $\beta_{d,1}$ | effect of crowding | ``w[1]`` |
| $\beta_{d,2}$ | crowding $\times$ size effect | ``w[2]`` |
| $\beta_{c,k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| *a* | first parameter for growth variance | ``tau`` |
| *b* | second parameter for growth variance | ``tauSize`` |

@@

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\beta_{0,t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
| $\beta_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\beta_{d,1}$ | effect of plot cover | ``dd`` |
| $\beta_{c,k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| *p* | mixing fraction for effective plot cover | ``u`` |
| $\zeta$ | size parameter for negative binimial likelihood | ``theta`` |

@@

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\beta_{0,t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
| $\beta_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\beta_{s,t}$ | cover effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_s}$ | global cover effect | ``b1_mu`` |
| $\beta_{c,k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| $\tau$ | model variance in log normal likelihood | ``tau`` |

@@

\begin{align}
y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}) \\
\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
\end{align}

where $x_{i,q,t}$ is the log of genet *i* basal area at time *t*, $\beta_{0,t}$ is a year specific intercept, $\beta_Q$ is the random effect of quadrat group location, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *p* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *p*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet at time *t* ($w_{i,q,t}$), and $\beta_{d,2}$ is a size by crowding ($x_{i,q,t}w_{i,q,t}$) interaction effect.

@@

\begin{align}
&\mathtt{data} \qquad &y_{i,q,t}^{S} \sim \text{Bernoulli}(s_{i,q,t}) \\
&\mathtt{process} \qquad &\text{logit}(s_{i,q,t}) = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2 \textbf{I})\\
& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2 \textbf{I}) \\
& \qquad & \boldsymbol{\beta_{Q}} \sim \text{Normal}(0, \sigma_Q^2 \textbf{I}) \\
& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
& \qquad & \beta_{d,1} \sim \text{Normal}(0, 100) \\
& \qquad & \beta_{d,2} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
& \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 5)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2 | y_{i,q,t}^{S}] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{i,q,t}^{S} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}] \times \\
&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\beta_c|\sigma_{beta_c}^2] \times \\
&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] \times \\
&[\beta_{d,1}] [\beta_{d,1}] [\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}]
\end{align}

**Stan code for model**
```{r survival_stan, echo=FALSE, eval=FALSE}
data{
  // All Data
  int<lower=0> N; // observations
  int<lower=0> Yrs; // years
  int<lower=0> yid[N]; // year id
  int<lower=0> Covs; // climate covariates
  int<lower=0> G; // groups
  int<lower=0> gid[N]; // group id
  int<lower=0,upper=1> Y[N]; // observation vector
  matrix[N,Covs] C; // climate matrix
  vector[N] X; // size vector
  matrix[N,2] W; // crowding matrix
  real beta_tau; // prior sdev for climate effects
}
parameters{
  real a_mu;
  vector[Yrs] a;
  real b1_mu;
  vector[Yrs] b1;
  vector[Covs] b2;
  vector[2] w;
  vector[G] gint;
  real<lower=0> sig_a;
  real<lower=0> sig_b1;
  real<lower=0> sig_G;
}
transformed parameters{
  real mu[N];
  vector[N] climEff;
  vector[N] crowdEff;
  climEff <- C*b2;
  crowdEff <- W*w;
  for(n in 1:N){
    mu[n] <- inv_logit(a[yid[n]] + gint[gid[n]] + 
                         b1[yid[n]]*X[n] + crowdEff[n] + climEff[n]);
  }
}
model{
  // Priors
  a_mu ~ normal(0,100);
  w ~ normal(0,100);
  b1_mu ~ normal(0,100);
  sig_a ~ cauchy(0,5);
  sig_b1 ~ cauchy(0,5);
  sig_G ~ cauchy(0,5);
  gint ~ normal(0, sig_G);
  b2 ~ normal(0, beta_tau);
  a ~ normal(a_mu, sig_a);
  b1 ~ normal(b1_mu, sig_b1);

@@

\begin{align}
y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}\textbf{I}) \\
\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
\end{align}

@@

**Data, process, and parameter models.**
\begin{align}
&\mathtt{data} \qquad &y_{i,q,t+1}^G \sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}\textbf{I}) \\
&\mathtt{process} \qquad &\mu_{i,q,t+1} = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
& \qquad & \sigma^2_{iQ,t} = ae^{b\mu_{iQ,t+1}} \\
&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2 \textbf{I})\\
& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2 \textbf{I}) \\
& \qquad & \boldsymbol{\beta_{Q}} \sim \text{Normal}(0, \sigma_Q^2 \textbf{I}) \\
& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
& \qquad & \beta_{d,1} \sim \text{Normal}(0, 100) \\
& \qquad & \beta_{d,2} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
& \qquad & \sigma_Q \sim \text{Cauchy}(0, 2) \\
& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 5) \\
& \qquad & a \sim \text{Normal}(0, 100) \\
& \qquad & b \sim \text{Normal}(0, 100)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2,a,b | y_{i,q,t+1}^G] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{i,q,t+1}^G | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, a, b] \times \\
&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\beta_c|\sigma_{beta_c}^2] \times \\
&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] \times \\
&[\beta_{d,1}] [\beta_{d,1}] [\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}] [a] [b]
\end{align}

@@

\begin{align}
\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right)
\end{align}

@@

**Data, process, and parameter models**
\begin{align}
&\mathtt{data} \qquad &y^{R}_{q,t+1} \sim \text{NegBin}(\lambda_{q,t+1},\phi) \\
&\mathtt{process} \qquad &\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right) \\
&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2 \textbf{I})\\
& \qquad & \beta_{Q} \sim \text{Normal}(0,\sigma_Q^2 \textbf{I}) \\
& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
& \qquad & \beta_{d} \sim \text{Uniform}(-10, 10) \\
& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
& \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
& \qquad & \phi \sim \text{Uniform}(0, 10) \\
& \qquad & u \sim \text{Uniform}(0, 1)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\beta_0}, \tilde{\beta}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d}, \sigma_Q^2, \sigma_{\beta_0}^2, \phi, u | y^{R}_{q,t+1}] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y^{R}_{q,t+1} | \beta_{0,t}, \beta_{Q}, \beta_d, \boldsymbol{\beta_c}, \phi, u] [\beta_{0,t} | \tilde{\beta}, \sigma_{\beta_0}^2] \times \\
&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] [\boldsymbol{\beta_c}] [\beta_d] [\tilde{\beta_0}] [\sigma_Q] [\sigma_{\beta_0}] [\phi] [u]
\end{align}

**Stan code for model**
```{r rec_stan, echo=FALSE, eval=FALSE}
data{
  int<lower=0> N; // observations
  int<lower=0> Yrs; // years
  int<lower=0> yid[N]; // year id
  int<lower=0> Covs; // climate covariates
  int<lower=0> G; // groups
  int<lower=0> gid[N]; // group id
  int<lower=0> Y[N]; // observation vector
  matrix[N,Covs] C; // climate matrix
  vector[N] parents1; // crowding vector
  vector[N] parents2; // crowding vector
  real<lower=0> tau; // prior variance
}
parameters{
  real a_mu;
  vector[Yrs] a;
  vector[Covs] b2;
  real dd;
  real gint[G];
  real<lower=0> sig_a;
  real<lower=0> theta;
  real<lower=0> sig_G;
  real<lower=0, upper=1> u;
}
transformed parameters{
real mu[N];
vector[N] climEff;
vector[N] trueP1;
vector[N] trueP2;
vector[N] lambda;
vector[N] q;
climEff <- C*b2;
  for(n in 1:N){
    trueP1[n] <- parents1[n]*u + parents2[n]*(1-u);
    trueP2[n] <- sqrt(trueP1[n]);
    mu[n] <- exp(a[yid[n]] + gint[gid[n]] + dd*trueP2[n] + climEff[n]);
    lambda[n] <- trueP1[n]*mu[n];
    q[n] <- lambda[n]*theta;
  }
}
model{
// Priors
  u ~ uniform(0,1);
  theta ~ uniform(0,100);
  a_mu ~ normal(0,1000);
  dd ~ uniform(-100,100);
  sig_a ~ cauchy(0,5);
  sig_G ~ cauchy(0,5);
  for(g in 1:G)
   gint[g] ~ normal(0, sig_G);
  for(y in 1:Yrs){
    a[y] ~ normal(a_mu, sig_a);
  }
  for(j in 1:Covs)
    b2[j] ~ normal(0, tau);

@@

\begin{align}
y_{q,t+1}^{P} &\sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1} \\
\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c
\end{align}

>>>>

##  Hierarchical Bayesian models for vital rates
Here we describe the statistical models in complete detail, including full description of the Bayesian models and Stan code for each model. We write each model as three parts: (1) a data model (likelihood), (2) a process model, and (3) a parameter model. We then combine these three models to write the posterior and joint distributions as a single model statement. Following the full model expression we provide the Stan code that directly corresponds the statement of posterior and joint distributions. Note that Stan uses standard deviations ($\sigma$) to specify distributions rather than variances ($\sigma^2$)  In our expressions of the posterior and joint distributions we use general probability notation where [.] represents some unspecified distribution. We do this for clarity of model presentation. All prior distributions are defined in the parameter model sections. For clarity, we avoid subscripting species.

@@

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\beta_{0,t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
| $\beta_{Q,q}$ | random effect of quadrat group *q* | ``gint[q]`` where ``q`` is the quadrat group |
| $\beta_{s,t}$ | size effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_s}$ | global size effect | ``b1_mu`` |
| $\beta_{d,1}$ | effect of crowding | ``w[1]`` |
| $\beta_{d,2}$ | crowding $\times$ size effect | ``w[2]`` |
| $\beta_{c,k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| *a* | first parameter for growth variance | ``tau`` |
| *b* | second parameter for growth variance | ``tauSize`` |

@@

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\beta_{0,t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
| $\beta_{Q,q}$ | random effect of quadrat group *q* | ``gint[q]`` where ``q`` is the quadrat group |
| $\beta_{d,1}$ | effect of plot cover | ``dd`` |
| $\beta_{c,k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| *p* | mixing fraction for effective plot cover | ``u`` |
| $\zeta$ | size parameter for negative binimial likelihood | ``theta`` |

@@

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\beta_{0,t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
| $\beta_{Q,q}$ | random effect of quadrat group *q* | ``gint[q]`` where ``q`` is the quadrat group |
| $\beta_{s,t}$ | cover effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_s}$ | global cover effect | ``b1_mu`` |
| $\beta_{c,k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| $\tau$ | model variance in log normal likelihood | ``tau`` |

@@

\begin{align}
y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}) \\
\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
\end{align}

where $x_{i,q,t}$ is the log of genet *i* basal area at time *t*, $\beta_{0,t}$ is a year specific intercept, $\beta_{Q,q}$ is the random effect for the *q*th quadrat to account for spatial variation, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *p* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *p*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet at time *t* ($w_{i,q,t}$), and $\beta_{d,2}$ is a size by crowding ($x_{i,q,t}w_{i,q,t}$) interaction effect.

@@

\begin{align}
& \mathtt{data} \qquad &y_{i,q,t}^{S} \sim \text{Bernoulli}(s_{i,q,t}) \\
& \mathtt{process} \qquad &\text{logit}(s_{i,q,t}) = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
& \mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2)\\
& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2) \\
& \qquad & \beta_{Q,q} \sim \text{Normal}(0, \sigma_Q^2) \\
& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(\textbf{0}, \sigma_c^2 \textbf{I}) \\
& \qquad & \beta_{d,1} \sim \text{Normal}(0, 100) \\
& \qquad & \beta_{d,2} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
& \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 5)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2 | y_{i,q,t}^{S}] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{i,q,t}^{S} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}] \times \\
&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
&\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] \times \\
&[\beta_{d,1}] [\beta_{d,1}] [\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}]
\end{align}

**Stan code for model**
```{r survival_stan, echo=TRUE, eval=FALSE}
data{
  // All Data
  int<lower=0> N; // observations
  int<lower=0> Yrs; // years
  int<lower=0> yid[N]; // year id
  int<lower=0> Covs; // climate covariates
  int<lower=0> G; // groups
  int<lower=0> gid[N]; // group id
  int<lower=0,upper=1> Y[N]; // observation vector
  matrix[N,Covs] C; // climate matrix
  vector[N] X; // size vector
  matrix[N,2] W; // crowding matrix
  real beta_tau; // prior sdev for climate effects
}
parameters{
  real a_mu;
  vector[Yrs] a;
  real b1_mu;
  vector[Yrs] b1;
  vector[Covs] b2;
  vector[2] w;
  vector[G] gint;
  real<lower=0> sig_a;
  real<lower=0> sig_b1;
  real<lower=0> sig_G;
}
transformed parameters{
  real mu[N];
  vector[N] climEff;
  vector[N] crowdEff;
  climEff <- C*b2;
  crowdEff <- W*w;
  for(n in 1:N){
    mu[n] <- inv_logit(a[yid[n]] + gint[gid[n]] + 
                         b1[yid[n]]*X[n] + crowdEff[n] + climEff[n]);
  }
}
model{
  // Priors
  a_mu ~ normal(0,10);
  w ~ normal(0,10);
  b1_mu ~ normal(0,10);
  sig_a ~ cauchy(0,5);
  sig_b1 ~ cauchy(0,5);
  sig_G ~ cauchy(0,5);
  gint ~ normal(0, sig_G);
  b2 ~ normal(0, beta_tau);
  a ~ normal(a_mu, sig_a);
  b1 ~ normal(b1_mu, sig_b1);

@@

\begin{align}
y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}) \\
\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
\end{align}

@@

**Data, process, and parameter models.**
\begin{align}
&\mathtt{data} \qquad &y_{i,q,t+1}^G \sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}) \\
&\mathtt{process} \qquad &\mu_{i,q,t+1} = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
& \qquad & \sigma^2_{iQ,t} = ae^{b\mu_{iQ,t+1}} \\
&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2)\\
& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2) \\
& \qquad & \boldsymbol{\beta_{Q,q}} \sim \text{Normal}(0, \sigma_Q^2) \\
& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(\textbf{0}, \sigma_c^2 \textbf{I}) \\
& \qquad & \beta_{d,1} \sim \text{Normal}(0, 100) \\
& \qquad & \beta_{d,2} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
& \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 5) \\
& \qquad & a \sim \text{Normal}(0, 100) \\
& \qquad & b \sim \text{Normal}(0, 100)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2,a,b | y_{i,q,t+1}^G] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{i,q,t+1}^G | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, a, b] \times \\
&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
&\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] \times \\
&[\beta_{d,1}] [\beta_{d,1}] [\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}] [a] [b]
\end{align}

@@

\begin{align}
\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right)
\end{align}

@@

**Data, process, and parameter models**
\begin{align}
&\mathtt{data} \qquad &y^{R}_{q,t+1} \sim \text{NegBin}(\lambda_{q,t+1},\phi) \\
&\mathtt{process} \qquad &\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right) \\
&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2)\\
& \qquad & \beta_{Q,q} \sim \text{Normal}(0,\sigma_Q^2) \\
& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(\textbf{0}, \sigma_c^2 \textbf{I}) \\
& \qquad & \beta_{d} \sim \text{Uniform}(-10, 10) \\
& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
& \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
& \qquad & \phi \sim \text{Uniform}(0, 10) \\
& \qquad & u \sim \text{Uniform}(0, 1)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\beta_0}, \tilde{\beta}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d}, \sigma_Q^2, \sigma_{\beta_0}^2, \phi, u | y^{R}_{q,t+1}] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y^{R}_{q,t+1} | \beta_{0,t}, \beta_{Q}, \beta_d, \boldsymbol{\beta_c}, \phi, u] \times \\
&[\beta_{0,t} | \tilde{\beta}, \sigma_{\beta_0}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
&\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] [\boldsymbol{\beta_c}] [\beta_d] [\tilde{\beta_0}] [\sigma_Q] [\sigma_{\beta_0}] [\phi] [u]
\end{align}

**Stan code for model**
```{r rec_stan, echo=TRUE, eval=FALSE}
data{
  int<lower=0> N; // observations
  int<lower=0> Yrs; // years
  int<lower=0> yid[N]; // year id
  int<lower=0> Covs; // climate covariates
  int<lower=0> G; // groups
  int<lower=0> gid[N]; // group id
  int<lower=0> Y[N]; // observation vector
  matrix[N,Covs] C; // climate matrix
  vector[N] parents1; // crowding vector
  vector[N] parents2; // crowding vector
  real<lower=0> tau; // prior variance
}
parameters{
  real a_mu;
  vector[Yrs] a;
  vector[Covs] b2;
  real dd;
  real gint[G];
  real<lower=0> sig_a;
  real<lower=0> theta;
  real<lower=0> sig_G;
  real<lower=0, upper=1> u;
}
transformed parameters{
real mu[N];
vector[N] climEff;
vector[N] trueP1;
vector[N] trueP2;
vector[N] lambda;
vector[N] q;
climEff <- C*b2;
  for(n in 1:N){
    trueP1[n] <- parents1[n]*u + parents2[n]*(1-u);
    trueP2[n] <- sqrt(trueP1[n]);
    mu[n] <- exp(a[yid[n]] + gint[gid[n]] + dd*trueP2[n] + climEff[n]);
    lambda[n] <- trueP1[n]*mu[n];
    q[n] <- lambda[n]*theta;
  }
}
model{
// Priors
  u ~ uniform(0,1);
  theta ~ uniform(0,10);
  a_mu ~ normal(0,10);
  dd ~ uniform(-10,10);
  sig_a ~ cauchy(0,5);
  sig_G ~ cauchy(0,5);
  for(g in 1:G)
   gint[g] ~ normal(0, sig_G);
  for(y in 1:Yrs){
    a[y] ~ normal(a_mu, sig_a);
  }
  for(j in 1:Covs)
    b2[j] ~ normal(0, tau);

@@

\begin{align}
y_{q,t+1}^{P} &\sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1} \\
\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c
\end{align}
